---
title: "Tidy Tuesday Feb. 10 2026 - 2026 Winter Olympics"
author: "Benjamin Sherman"
format: html
---
## Introduction

Today, we're going to look at data from the little brother to the Summer Olympics - the Winter Olympics. Let's load packages and data!

```{r}
#| label: package-preperation
#install.packages("tidytuesdayR")
library(tidytuesdayR)
library(tidyverse)
library(janitor)
```


```{r}
#| label: loadANDcreate-DATA
tuesdata <- tt_load('2026-02-10')

schedule <- tuesdata$schedule

glimpse(schedule)

#write_csv(schedule, "schedule.csv")
```

## Research Topic and Data Preperation

As you can see in the glimpse for the `schedule` data, the `event_description` gives us insight into the variety of schedule items. I want to focus on one specific observation of mine - the existence of olympic training events vs. actual olympic competitions. With the end goal of creating a visualization that offers insight into this trend, let's mutate the data and create a new variable (i.e. `training`) that is labeled TRUE if the event has the word "training" in its event description. Otherwise, it will be FALSE.

```{r}
#| label: training-variable-creation
schedule <- schedule |>
  mutate(
    is_training_event = if_else(
      str_detect(str_to_lower(event_description), "training"),
      TRUE,
      FALSE
    )
  )
```

More specifically, I'm curious which disciplines have a larger or smaller proportion of training events than others. Before making a visualization, let's get a glimpse of the overall trend.
```{r}
#| label: visualize-training
discipline_summary <- schedule |>
  group_by(discipline_code, discipline_name) |>
  summarise(
    total        = n(),
    training     = sum(is_training_event),
    non_training = sum(!is_training_event),
    pct_training = training / total,
    .groups = "drop"
  ) |>
  arrange(desc(pct_training))

discipline_summary
```

Now, before moving on, I'm going to double-check that my methodology of identifying training events through the existence of "training" in the `event_description` variable is flawed. I'm picking some disciplines that had no training events, and looking at the event descriptions for said disciplines.
```{r}
#| label: checking-methodology
no_training_disciplines <- c("FSK", "IHO", "SBD", "SMT", "SSK", "STK")
schedule |>
  filter(discipline_code %in% no_training_disciplines) |>
  count(discipline_name, event_description) |>
  arrange(discipline_name) |>
  print(n = Inf)
```

As can be seen, all of these events are actual competitions. My methodology seems sound. Now, let's visualize the relationship between disciplines and proportion of training vs. competition events!

## Visualization

```{r}
#| label: visualization
discipline_summary |>
  pivot_longer(cols = c(training, non_training),
               names_to  = "type",
               values_to = "count") |>
  mutate(
    type = factor(type, levels = c("non_training", "training"),
                  labels = c("Event", "Training")),
    discipline_name = fct_reorder(discipline_name, pct_training)
  ) |>
  ggplot(aes(x = discipline_name, y = count, fill = type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("Event" = "#2166ac", "Training" = "#f4a582")) +
  coord_flip() +
  labs(
    title    = "Proportion of Training vs. Events by Discipline",
    subtitle = "2026 Milan-Cortina Winter Olympics",
    x        = "Discipline Name",
    y        = "Proportion of Sessions",
    fill     = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

